---
title: "ADA Final Project"
author: "MARIAM BALOGUN"
date: "2024-11-23"
output: pdf_document
---

#TITLE OF PROJECT
The Association between Educational Attainment and Chronic Conditions (like Diabetes and Hypertension) among Medicare Beneficiaries: a cross-sectional study.

-Objectives of Study
a.	Primary: To determine if education level is associated with the prevalence of diabetes and hypertension among 2022 Medicare beneficiaries
b.	Secondary: To determine if this association differs by gender and across races

-Description of dataset & Study population
Data used is the Medicare Health Outcomes Survey (HOS) Research Data File, year 2022 Baseline Cohort. Data is publicly available at: https://www.hosonline.org/en/data-dissemination/research-data-files/
A random sample of 1,013,279 beneficiaries, including the aged and disabled. Of those sampled, 280,261 (27.7%) completed the survey and were included in the CSV data file

# Install packages and open libraries
```{r}
pacman::p_load(ggplot2, car, odds.n.ends, readr, dplyr, tidyverse, table1, DiagrammeR, rsvg)
```

# Import data for this project
```{r}
#Import file
FinalProjectData <- read.csv("/Users/mac/Desktop/MASTERS AT BROWN SCHOOL/FALL 2024 - ADVANCED DATA ANALYTICS/Medicare HOS DATA.csv")

#view file
dim(FinalProjectData) #Dataset has 81 variables and 280261 observations

```

# Data Management
```{r}
# Check the dataset's structure and summary
str(FinalProjectData)
summary(FinalProjectData)

#Rename the diabetes and hypertension status column names
FinalProjectData <- FinalProjectData %>%
  rename(
    diab_stat = B25ccdiabet,     #'B25ccdiabet' to 'diab_stat'
    hyprtnsn_stat = B25cchbp)    #'B25cchbp' to 'hyprtnsn_stat')

# Check the updated column names
colnames(FinalProjectData)

# Recode variables into categorical factors
FinalProjectData <- FinalProjectData %>%
  mutate(
    EDUC = factor(EDUC, 
                  levels = c(1, 2, 3), 
                  labels = c("Less than HS", "HS/GED", "More than HS")),
    RACE = factor(RACE, 
                  levels = c(1, 2, 3), 
                  labels = c("White", "Black or African American", "Other")),
    GENDER = factor(GENDER, 
                    levels = c(1, 2), 
                    labels = c("Male", "Female")),
    diab_stat = factor(diab_stat, 
                       levels = c(1, 2), 
                       labels = c("Yes", "No")),
    hyprtnsn_stat = factor(hyprtnsn_stat, 
                           levels = c(1, 2), 
                           labels = c("Yes", "No")),
    MARITL = factor(MARITL, 
                    levels = c(1, 2), 
                    labels = c("Married", "Non-Married")),
    AGE = factor(AGE,
                 levels = c(1, 2, 3),
                 labels = c("Under 65", "65-74", "75 and older")))


# Select only the needed variables for the project from the dataset
FinalProjectData <- FinalProjectData %>%
  select(CASE_ID, AGE, RACE, GENDER, EDUC, MARITL, diab_stat, hyprtnsn_stat)

```


# Handling missing data
```{r, eval = FALSE}
#Since this project is assuming that data are missing completely at random, we would do a complete case analysis to remove missing data from each of the variables

#summary of each variable in the dataframe to determine how much missing data there is 
summary(FinalProjectData)

#Drop missing data from each variable
FinalProjectData1 <- FinalProjectData %>%
  drop_na(diab_stat)   #260739 observations left after dropping NAs in diab_stat

FinalProjectData2 <- FinalProjectData1 %>%
 drop_na(hyprtnsn_stat) #259038 observations left after dropping NAs in hyprtnsn_stat

FinalProjectData3 <- FinalProjectData2 %>%
  drop_na(EDUC)  #249120 observations left after dropping NAs from EDUC

FinalProjectData4 <- FinalProjectData3 %>%
  drop_na(CASE_ID, AGE, RACE, GENDER,MARITL) #drop NAs in other variables

view(FinalProjectData4) #Dataset now has 234,871 observations

```


#Create a Figure 1 Diagram showing a flowchart of how sample size was gotten
```{r}
grViz(diagram = "digraph flowchart{

      node [fontname = Helvetica, shape = rectangle, fontsize=15] 
      
      node1 [label = '@@1'] # starting number
      node2 [label = '@@2'] # number after exclusion 1
      node3 [label = '@@3'] # number after exclusion 2
      node4 [label = '@@4'] # number after exclusion 3
      node5 [label = '@@5'] # number after exclusion 4
      
      node1 -> node2 -> node3 -> node4 -> node5
}
      [1]: 'Number of paricipants in dataset n = 280,261'
      [2]: 'Excluding 19,522 with NAs on diabetes status variable n = 260,739'
      [3]: 'Excluding 1,701 with NAs data on hypertension status variable n = 259,038'
      [4]: 'Excluding 9,918 with NAs on educational level variable n = 249,120'
     [5]: 'Excluding 14,249 with NAs on other variables n = 234,871' ")
```

# Create a Table 1 that show demographics of the participants
```{r}
#Since this project is exploring two outcome variables (diabetes and hypertension status) in the participants, we shall present a Table 1 showing the demographics of participants with these combined outcomes as a way of exploring how the co-occurrence of the conditions impacts demographics.

# Create combined outcome variable
FinalProjectData4 <- FinalProjectData4 %>%
  mutate(
    combined_outcome = case_when(
      diab_stat == "Yes" & hyprtnsn_stat == "Yes" ~ "Both",
      diab_stat == "Yes" & hyprtnsn_stat == "No" ~ "Diabetes Only",
      diab_stat == "No" & hyprtnsn_stat == "Yes" ~ "Hypertension Only",
      diab_stat == "No" & hyprtnsn_stat == "No" ~ "Neither"
    ) %>%
      factor(levels = c("Both", "Diabetes Only", "Hypertension Only", "Neither")))

# Create Table 1 for combined outcomes
knitr::kable(table1(~ AGE + RACE + GENDER + EDUC + MARITL |combined_outcome, FinalProjectData4)) 

```


# Figure 2: Descriptives Bar Chart of variables
```{r}
# Summarize data for plotting
plot_data <- FinalProjectData4 %>%
  group_by(EDUC) %>%
  summarize(diabetes_count = sum(diab_stat == "Yes"))

plot_data2 <- FinalProjectData4 %>%
  group_by(EDUC) %>%
  summarize(hypertension_count = sum(hyprtnsn_stat == "Yes"))



# Generating the car charts for each outcome
ggplot(plot_data, aes(x = EDUC, y = diabetes_count, fill = EDUC)) +
  geom_bar(stat = "identity") +
  labs(title = "Number of those with Diabetes grouped by Education Level",
       x = "Education Level",
       y = "Count") + theme_minimal()

ggplot(plot_data2, aes(x = EDUC, y = hypertension_count, fill = EDUC)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Number of those with Hypertension grouped by Education Level",
    x = "Education Level",
    y = "Count" ) + theme_minimal() 


```


# Testing for assumptions: Multicollinearity
```{r}
# We will use the Variance Inflation Factor (VIF) to ensure that no multicollinearity exists among our predictors.

model_hyp <- glm(hyprtnsn_stat~ EDUC + AGE + RACE + GENDER + MARITL, 
             data = FinalProjectData4, 
             family = binomial)

model_diab <- glm(diab_stat~ EDUC + AGE + RACE + GENDER + MARITL, 
             data = FinalProjectData4, 
             family = binomial)

# Calculate VIF
vif(model_hyp)
vif(model_diab)

#RESULT: For each of the outcome variables (diab_stat and hyprtnsn_stat), their models show no multicollinearity exists among the predictors
```
RESULT: 
For each of the outcome variables (diab_stat and hyprtnsn_stat), their models show no multicollinearity exists among the predictors

#Fitting Logistics Regression model for the outcomes

Hypotheses
• H0: The probability of having Chronic conditions like Diabetes or hypertension is not associated with Educational level.
• Ha: The probability of having Chronic conditions like Diabetes or hypertension is associated with Educational level.

```{r}
# Fitting logistic regression model for diabetes status among participants
diabetes_model <- glm(diab_stat ~ EDUC + AGE + RACE + GENDER, 
                      data = FinalProjectData4, 
                      family = binomial)

# Summary of results
summary(diabetes_model)
broom::tidy(diabetes_model, exponentiate = TRUE, conf.int = TRUE)

```
-Interpretation
For those with 'High school education', the odds of having Diabetes is 1.32 times those with 'less than a high school education', while those in the 'more than a high school education' have 1.69 times higher odds than those with 'less than a high school education'

Since the confidence interval does not include 1, the association is statistically significant at the 95% confidence level. This suggests that the true OR is between 1.56 to 1.64

```{r}
# Fitting logistic regression model for hypertension status 
hypertension_model <- glm(hyprtnsn_stat ~ EDUC + AGE + RACE + GENDER, 
                          data = FinalProjectData4, 
                          family = binomial)

#Summary of results
summary(hypertension_model)
broom::tidy(hypertension_model, exponentiate = TRUE,  conf.int = TRUE)

```
-Interpretation
The odds of having Hypertension among those in the category of 'High school education' is 1.21 times greater than those with 
'less than a high school education', while those in the 'more than high school education' have 1.54 times greater odds than those with 'less than a high school education'.

Since the confidence interval does not include 1, the association is statistically significant at the 95% confidence level. This suggests that the true OR is between 1.50 to 1.58

#Testing effect modification by race and gender for the outcomes
```{r}
# Model with interaction terms for diabetes status
diabetes_model2 <- glm(diab_stat ~ EDUC * RACE + AGE + EDUC * GENDER, 
                      data = FinalProjectData4, 
                      family = binomial)

# Summary of the model
summary(diabetes_model2)
broom::tidy(diabetes_model2, exponentiate = TRUE, conf.int = TRUE)
```
-Interpretations
For Black or African Americans with a 'High school education', the odds of having DIABETES is O.86 times that of whites with 'less than a high school education'; while Black or African Americans with 'More than a High School education' have 0.67 times the odds of having DIABETES than whites with 'less than a high school education'. 

For those in the 'Other' race category with a 'High school education', the odds of having DIABETES is 0.85 times that of whites with 'less than a high school education'; while those in the 'Other' race category with 'More than a High School education' have 0.68 times the odds of having DIABETES than whites with with 'less than a high school education'. 

For females with a 'High school education', the odds of having DIABETES is 1.20 times that of Males with 'less than a high school education'; while females with 'More than a High School education' have 1.33 times the odds of having DIABETES than Males with 'less than a high school education'

With pvalues less than 0.05 for each group, it shows that race and gender significantly modifies the association between educational level and Diabetes status. 


```{r}
# Model with interaction terms for hypertension status
hypertension_model2 <- glm(hyprtnsn_stat ~ EDUC * RACE + AGE + EDUC * GENDER, 
                          data = FinalProjectData4, 
                          family = binomial)

# Summary of the model
summary(hypertension_model2)
broom::tidy(hypertension_model2, exponentiate = TRUE, conf.int = TRUE)
```

For Black or African Americans with a 'High school education', the odds of having HYPERTENSION is 1.04 times that of whites with 'less than a high school education'; while Black or African Americans with 'More than a High School education' have 0.81 times the odds of having Diabetes than whites with 'less than a high school education'. 

For those in the 'Other' race category with a 'High school education', the odds of having HYPERTENSION is 0.92 times that of whites with 'less than a high school education'; while those in the 'Other' race category with 'More than a High School education' have 0.82 times the odds of having HYPERTENSION than whites with with 'less than a high school education'. 

For females with a 'High school education', the odds of having HYPERTENSION is 1.20 times that of Males with 'less than a high school education'; while females with 'More than a High School education' have 1.41 times the odds of having HYPERTENSION than Males with 'less than a high school education'

With pvalues less than 0.05 for each group, it shows that race and gender significantly modifies the association between educational level hypertension status. 

#Perform Likelihood ratio tests to compare models with and without the interaction terms to formally test for effect modification.
```{r}
# Perform likelihood ratio test
anova(diabetes_model, diabetes_model2, test = "LRT")
```
-Interpretation
The significant p-value which is < 0.05 indicates that adding interaction terms for race and gender significantly improves the model fit. That is, race and gender modify the relationship between education level and diabetes status.

```{r}
# Perform likelihood ratio test
anova(hypertension_model, hypertension_model2, test = "LRT")
```

-Intepretation
The significant p-value which is < 0.05 indicates that adding interaction terms for race and gender significantly improves the model fit. That is, race and gender modify the relationship between education level and hypertension status.



Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.



